Paper 4: Praise the people or praise the place?

### 2024-01-17 paper.qmd
change spell checker

To answer this, the investigation leverages two novel and rich data sources. The first, "Vem är Vem", is a comprehensive set of biographical dictionaries containing the profiles of 75,000 notable Swedes active between 1945 and 1968. The second, the "Svensk Industrikalender" or Swedish Industrial Calendar of 1947, offers an exhaustive catalogue of industrial firms, detailing their activities, workforce, and financial metrics.

@@

A number of recent papers have made use of biographic data in innovative ways. @ford2023not use biographies of high school graduates compiled at the time of their school reunions to create a far richer measure of human capital than the conventional measure, number of years of schooling, alone. These biographies are "mini-CVs", containing information about the school leaver's grades, their occupational trajectory, and their family background. These are used to create an innovative approach to measuring `upper tail knowledge`.

@Nekoei2020Herstory titled "Herstory: the rise of self-made women" analyzes the historical prominence of self-made women using a specially created database. This database, formed by applying machine learning to Wikidata and Wikipedia, catalogues notable individuals throughout history, highlighting details like occupation and family background. Their unique approach reveals a significant increase in the number of prominent women, especially those who achieved success independently of their family connections, across various fields since the 17th century. This research provides a fresh perspective on women's historical achievements and roles.

Importantly, these papers go beyond the use of just administrative records, though it is a noble task.

In this paper, I structure biographic data about elite individuals that is similar in structure to @ford2023not, and use the career trajectories of these individuals to better understand the contribution of educated workers to the adoption of electricity across Sweden in the 20th century. The contribution is the scale - I capture 75,000 individuals in Sweden in an automated manner.

@@

Goldin's extensive analysis of labor markets in the 20th century provides a comprehensive backdrop to this study [-@Goldin1994; -@Goldin1998]. Her work highlights critical shifts in labor participation, wage structures, and job security, reflecting the complex interplay between societal changes and labor market dynamics. These insights are crucial for understanding how shifts in human capital, like those during Sweden's electrification period, contribute to broader economic outcomes.

@@

'Big data' papers in economic history now abound, as surveyed in @gutmann2018big in a review titled "'Big Data' in Economic History". Many of these papers construct and use high quality register or census-like data on individual outcomes. Notable examples include the Longitudinal, Intergenerational Family Electronic Micro-Database Project by Martha Bailey which focuses on family histories to understand long-term economic trends. These histories are collected from various census-like sources and innovative ML tools are used to construct the longitudinal links. Similarly, "The Making of Modern America: Migratory Flows in the Age of Mass Migration" by @bandiera2013making involved the digitization of 24 million records of migrant flows through Ellis Island in New York, and found that measured out-migration rates in the US were double the reported figures in the earliest decades of the 20th century. @eichengreen2021gold uses natural language processing (NLP) to understand the content of a parliamentary committee debate on the gold standard in South Africa.

There is also a growing literature of papers which lay out the step by step process required for economic historians to make use of these new tools. A great paper in this vein that bridges the gap between cutting-edge computer science literature and the use cases of applied economists is by @correia2023digitizing. The paper "Digitizing Historical Balance Sheet Data: A Practitioner’s Guide" explores the application of machine learning, particularly Optical Character Recognition (OCR), to digitize large-scale historical economic data. The authors highlight the limitations of off-the-shelf OCR software, mainly due to high error rates, and propose a combination of pre- and post-processing methods to enhance accuracy. They apply these methods to two extensive datasets of balance sheets and introduce "quipucamayoc," a Python package that unifies these techniques.

@amujala2023digitization explain how we can digitize bank records

melissa dell.

@@

"Vem är Vem?" is a biographical dictionary, comprising a rich repository of information about notable individuals in Sweden. Published in two regional editions with a total of five volumes each, the first edition spanned from 1945 to 1950, and the second from 1962 to 1968, by the Bokförlaget Vem är Vem publishing house. An additional volume specifically focussed on individuals in industry and business was produced in 1945. This encyclopedia offers an invaluable snapshot of Swedish societal and professional landscapes during these pivotal periods. [fix citation]

The primary intention behind the creation of "Vem är Vem?" was to spotlight individuals who were at the peak of their careers, regardless of their age. This focus extends beyond traditional measures of influence, emphasizing the importance of those in influential positions or notable roles across relatively diverse sectors. As such, it serves as a crucial resource for understanding the professional and personal trajectories of around 75,000 individuals who shaped Swedish society in the mid-20th century.

@@

"Vem är Vem?" is useful to economic historians thanks to its high quality digitization, with nine out of the 11 total volumes being made accessible online by librarians in Uppsala through *Projekt Runeberg*, as shown in @fig-who-is-who-books. This digitization has facilitated research, allowing for a broader exploration of the biographies and career paths of thousands of individuals. The encyclopedia's extensive coverage makes it a goldmine for researchers, historians, and anyone interested in the socio-economic history of Sweden during a period marked by significant change and development.

@@

In the context of economic and historical research, "Vem är Vem?" serves as a unique tool. By providing detailed biographies and career information, it allows for an in-depth analysis of the human capital that contributed to Sweden's economic and social evolution during the mid-20th century.

@@

@@

The Svensk Industrikalender from 1947, published by Sveriges Industriförbund (Sweden's Industrial Association), is a comprehensive directory of Swedish industrial firms. This calendar was issued annually from 1918 to 2000 and contains information related to Swedish industry. The 1947 edition available on the Project Runeberg website was digitized in April 2012, sourced from the Centrum för Näringslivshistoria. The calendar is believed to be under catalog protection but not copyright [citation].

@@

@@

In order to analyze both the biographical dictionaries and industrial catalogue, we need to bend the text into a machine readable structure. This process is relatively complicated. It involves breaking each component of the source up (e.g. each biography or company record), extracting the pertinent information from each record, storing each value with its associated key, and then saving this information in a way that is easy to analyze and aggregate.

@@

Prior to the advent of Large Language Models (LLMs), this was a task that required a large number of human hours to complete, either putting the information into an excel sheet by hand, or writing rules to extract the information from the text. The first approach limits the number of observations a researcher can collect on her own, and the second approach quickly turns into the first.

Due to the number of abbreviations, acronyms and contractions (for example, `Gävleb. l.` is the contraction of *Gävleborg län* or Gävleborg county in @fig-who-is-who-page), while it might be possible to take a simple rules based approach to replacing these contractions with their complete Swedish text, and then looking for specific terms relating to each piece of information, the number of rules soon balloons to an unreasonable figure. Writing a rule for every case necessitates as much human involvement as would be required to manually structure the information - the first approach.

However, with the rapid advancements in LLM technology in the previous five years, and poplular adoption of these tools through Chat-GPT and Microsoft's integration of GPTs into their products in the previous year, new tools mean this manual workload can be avoided to a large extent.

I make use of the backed of Chat-GPT, a model called GPT-3.5-turbo from OpenAI to structure the information from the dictionaries and catalogue into a JSON format that I can analyse, step 3, as shown in @fig-data-collection-process.

@@

The GPT-3.5-turbo model which I make use of is a LLM which has been trained on all of Wikipedia and Wikidata, among other training material. These two sources contain the same information, but in a different format, as shown in the adapted extracts below. As the base model that is pre-trained to predict the next token on this kind of data, it has developed the ability though repeated exposure to this kind of biographic data to produce structured information from free text, and likewise construct natural sounding text from pieces of structured information.

@@

```{r}
library(tibble)

@@

wenstrom_data %>% 
gt() %>% 
cols_label(key = "Key", value = "Value") %>% 
tab_header(title = md("Wikidata information about Swedish inventor **Jonas Wenström**"))

@@

@@

Following this process of structuring the records into a format with specified keys and values, I augment the data by geocoding locations in order to analyse geographic paths of individuals in the sample, and geographic clusters of firms.

@@

person_info %>% 
gt() %>% 
cols_label(key = "Key", value = "Value") %>% 
tab_header(title = md("Selected collected information about Swedish engineer and power station manager<br>**Axel Verner Nordell**"))

@@

The next task involved grouping the firms from the catalogue and then classifying each individual in the biographical dictionaries into a particular occupation in order to track the trends in the labour market.

For both of these tasks, I lent on the tools of text embeddings, and a combination of unsupervised machine learning and advanced language processing techniques. I made use of a text embedding model trained on Swedish language text by the Royal Library @bert-base-swedish-cased. It is an adaptation of the breakthrough BERT model, introduced by Google Research in 2018 [citation]. The advantage of this model is that it has been trained on a selection of Swedish data, including books, news reports, and internet forums. Hence it is able to score the similarity of Swedish business descriptions and occupational titles.

Text embeddings are effective for clustering because they capture semantic meaning rather than relying on surface-level features like character composition. For example, while "steam engine" and "power station" are different in characters and literal meaning, they are semantically related in the context of industrial machinery and energy production. Text embeddings transform these phrases into numerical vectors that reflect this semantic similarity. When applied to clustering, this means that items with similar meanings, even if their literal expressions differ, are grouped together based on the contextual and conceptual similarities encoded in their embeddings. This capability makes text embeddings particularly powerful for organizing and categorizing text data in a way that aligns with human understanding and interpretation. [citation]

For grouping firms, I used the k-means clustering algorithm, an unsupervised learning method, to categorize firms into 20 distinct categories. This algorithm works by partitioning the data into k distinct clusters based on features (in this case, text embeddings derived from company descriptions). [citation]

@@

In classifying occupations, I utilized the KB BERT model to create sentence embeddings for both the titles in the three-digit Historical International Standard Classification of Occupations (HISCO) schema and for each occupation in my biographical dictionary data. These embeddings were then projected into a 364-dimensional vector space. For each occupation, I determined the closest HISCO code in this vector space based on cosine distance, a measure of similarity without a specific unit. I established a threshold for this distance to ensure that the occupation was 'close enough' to the corresponding HISCO code, setting it to a level where 85% of occupations received a HISCO code. While somewhat arbitrary, this approach allowed for a contextually relevant classification of occupations, drawing on the advanced language understanding capabilities of the KB BERT model.

## What does the industrial calendar tell us about industrial firms in Sweden in 1947?

@@

[In this map, we show two kinds of firms; firms taht make electronic products, and firms that produce power or supply power stations with equipment. We can see that electronic products companies are clustered near the large population centers of Stockholm, Gothenburg and Malmö. There are many power stations in central Southern Sweden that make use of hydropower, and firms near them that supply equipment for these power stations.]

@@

>>>>

To answer this, the investigation leverages two novel and rich data sources. The first, *Vem är Vem*, is a comprehensive set of biographical dictionaries containing the profiles of 75,000 notable Swedes active between 1945 and 1968. The second, the "Svensk Industrikalender" or Swedish Industrial Calendar of 1947, offers an exhaustive catalogue of industrial firms, detailing their activities, workforce, and financial metrics.

@@

Several recent papers have made use of biographic data in innovative ways. @ford2023not use biographies of high school graduates compiled at the time of their school reunions to create a far richer measure of human capital than the conventional measure, number of years of schooling, alone. Titled, "Not the Best Fillers in of Forms? The Danish and Norwegian Graduate Biographies and 'Upper Tail Knowledge'", the authors explain that these biographies are "mini-CVs", containing information about the school leaver's grades, their occupational trajectory, and their family background. These are used to create an innovative approach to measuring `upper tail knowledge`.

@Nekoei2020Herstory titled "Herstory: the rise of self-made women" analyzes the historical prominence of self-made women using a specially created database. This database, formed by applying machine learning to Wikidata and Wikipedia, catalogues notable individuals throughout history, highlighting details like occupation and family background. Their unique approach reveals a significant increase in the number of prominent women, especially those who achieved success independently of their family connections, across various fields, starting with literature, since the 17th century. This research provides a fresh perspective on women's historical achievements and roles.

Importantly, these papers go beyond the use of just administrative records, which contain register-like data that economic historians are familiar with. Leveraging new kinds of sources in this way allows the authors to approach their research with different kinds of questions.

In this paper, I structure biographic data about elite individuals that is similar in structure to @ford2023not, and use the career trajectories of these individuals to better understand the contribution of educated workers to the adoption of electricity across Sweden in the 20th century. The differentiator in this case is the scale - I capture 75,000 individuals in Sweden in an automated manner, or about one percent of the population of the country at the time.

@@

Claudia Goldin's extensive analysis of labor markets in the 20th century provides a comprehensive backdrop to this study [-@Goldin1994; -@Goldin1998]. Her work highlights critical shifts in labor participation, wage structures, and job security, reflecting the complex interplay between societal changes and labor market dynamics. These insights are crucial for understanding how shifts in human capital, like those during Sweden's electrification period, contribute to broader economic outcomes.

@@

'Big data' papers in economic history now abound, as surveyed in @gutmann2018big in a review titled *'Big Data' in Economic History*. Many of these papers construct and use high quality register or census-like data on individual outcomes. Notable examples include the Longitudinal, Intergenerational Family Electronic Micro-Database Project by Martha Bailey which focuses on family histories to understand long-term economic trends. These histories are collected from various census-like sources and innovative ML tools are used to construct the longitudinal links. Similarly, "The Making of Modern America: Migratory Flows in the Age of Mass Migration" by @bandiera2013making involved the digitization of 24 million records of migrant flows through Ellis Island in New York, and found that measured out-migration rates in the US were double the reported figures in the earliest decades of the 20th century. @eichengreen2021gold uses natural language processing (NLP) to understand the content of a parliamentary committee debate on the gold standard in South Africa. Clark and Cummins' families of England database contains 1.7m marriage records in England from the 19th to the 21st centuries and allows to authors to pry out social dynamics in family formation, as well as geographic sorting between the North and South of England [-@Clark2018BigSort].

There is also a growing literature in which authors lay out the step by step processes required for economic historians to make use of these new tools. A great paper in this vein that bridges the gap between cutting-edge computer science literature and the use cases of applied economists is by @correia2023digitizing. The paper "Digitizing Historical Balance Sheet Data: A Practitioner’s Guide" explores the application of machine learning, particularly Optical Character Recognition (OCR), to digitize large-scale historical economic data. The authors highlight the limitations of off-the-shelf OCR software, mainly due to high error rates, and propose a combination of pre- and post-processing methods to enhance accuracy. They apply these methods to two extensive datasets of balance sheets and introduce "quipucamayoc," a Python package that unifies these techniques.

@amujala2023digitization, in "Digitization and data frames for card index records" explain the entire process through which they digitize and structure loan records from bank cards that contain both machine written and hand written text in varying formats over time. Tha authors lower the barrier to entry for other researchers by explaining their use off the shelf technology from Amazon Web Services. Each step is explained along with tips for successful extraction of handwritten information.

Perhaps the most impressive of these kinds of papers is the *Layout Parser* from Melissa Dell and her lab. The team have produced a python library that can be fine tuned to parse document type, extract information, and use machine learning to correct common errors at the point of data extraction. Dell demonstrates the use of this tool by extracting firm performance data from Japanese reports on yearly firm output which are atypical from the kinds of tables or column text that off the shelf optical character recognition tools have been trained on [-@dell2020large].

I hope that this paper can be useful for other researchers using similar kinds of sources as a prompt on where to start collecting and structuring their data at scale.

@@

*Vem är Vem?* is a biographical dictionary, comprising a rich repository of information about notable individuals in Sweden. Published in two regional editions with a total of five volumes each, the first edition spanned from 1945 to 1950, and the second from 1962 to 1968, by the Bokförlaget Vem är Vem publishing house [@harnesk1945vem]. An additional volume specifically focussed on individuals in industry and business was produced in 1945. This encyclopedia offers an invaluable snapshot of Swedish societal and professional landscapes during these pivotal periods.

The primary intention behind the creation of *Vem är Vem?* was to spotlight individuals who were at the peak of their careers, regardless of their age. This focus extends beyond traditional measures of influence, emphasizing the importance of those in influential positions or notable roles across relatively diverse sectors. As such, it serves as a crucial resource for understanding the professional and personal trajectories of around 75,000 individuals who shaped Swedish society in the mid-20th century.

@@

*Vem är Vem?* is useful to economic historians thanks to its high quality digitization, with nine out of the 11 total volumes being made accessible online by librarians in Uppsala through *Projekt Runeberg*, as shown in @fig-who-is-who-books. This digitization has facilitated research, allowing for a broader exploration of the biographies and career paths of thousands of individuals. The encyclopedia's extensive coverage makes it a goldmine for researchers, historians, and anyone interested in the socio-economic history of Sweden during a period marked by significant change and development.

@@

In the context of economic and historical research, 
*Vem är Vem?* serves as a unique tool. By providing detailed biographies and career information, it allows for an in-depth analysis of the human capital that contributed to Sweden's economic and social evolution during the mid-20th century.

@@

I use this biographical data to determine who the high skilled workers in electricity related jobs in Sweden were, and understand more about their career trajectories.

@@

The *Svensk Industrikalender*, published by Sveriges Industriförbund (Sweden's Industrial Association), is a comprehensive directory of Swedish industrial firms. This calendar was issued annually from 1918 to 2000 and contains information related to Swedish industry. The 1947 edition available on the *Project Runeberg* website was digitized in April 2012, sourced from the Centrum för Näringslivshistoria. The calendar is believed to be under catalog protection but not copyright .

@@

I use the industry calendar to uncover geographic clusters of firms that produced goods related to electricity for sale to consumers or to other businesses. I use these geographic clusters to understand the migration patterns among engineers in Sweden.

@@

In order to analyze both the biographical dictionaries and industrial catalogue, we need to bend the text into a machine readable structure. This process is not complicated, but somewhat involved. It includes breaking each component of the source up (e.g. each biography or company record), extracting the pertinent information from each record, storing each value with its associated key, and then saving this information in a way that is easy to analyze and aggregate.

@@

Prior to the advent of Large Language Models (LLMs), this structuring of data from free text into key-value pairs was a task that required a large number of human hours to complete. It could be done either by putting the information into an excel sheet by hand, or writing rules to extract the information from the text. The first approach limits the number of observations a researcher can collect on her own, and the second approach quickly turns into the first.

The biographical dictionaries are written in a specific way, with many abbreviations and contracted names for common field titles and values. Due to the number of abbreviations, acronyms and contractions (for example, `Gävleb. l.` is the contraction of *Gävleborg län* or Gävleborg county in @fig-who-is-who-page), while it might be possible to take a simple rules based approach to replacing these contractions with their complete Swedish text, and then looking with regular expressions for specific terms relating to each piece of information, the number of rules soon balloons to an unreasonable figure, making the process unwieldy at best and impossible at worst. Writing a rule for every case necessitates as much human involvement as would be required to manually structure the information - the first approach.

However, with the rapid advancements in LLM technology in the previous five years, and popular adoption of these tools through Chat-GPT and Microsoft's integration of GPTs into their products in the previous year, new tools mean this manual workload can be avoided to a large extent.

The structuring step involves sending a specific prompt to a LLM, along with the source material as text (rather than a scan of the book), and receiving back from the LLM a structured file with a key and value for each piece of information that I am interested in.

I make use of the computational backed of Chat-GPT, a model called GPT-3.5-turbo from OpenAI to structure the information from the dictionaries and catalogue into a JSON format that I can analyze, step 3, as shown in @fig-data-collection-process. Many other LLMs, some open source, are available. I have chosen GPT-3.5-turbo as it is simple to interact with it in the programming language Python, and because doing so is relatively affordable for contained workloads such as this, when compared to hosting such a model on your own, beefy, costly computer.

@@

The GPT-3.5-turbo model which I make use of is a LLM which has been trained on all of Wikipedia and Wikidata, among other training material. These two sources contain the same information, but in a different format, as shown in the adapted extracts below; the text and Table @tbl-wiki mimic the kind of material that the model I use is trained on.

As the base model underlying GPT-3.5-turbo is pre-trained to predict the next token on this kind of data, it has developed the ability though repeated exposure to this kind of biographic data to produce structured information from free text, and likewise construct natural sounding text from pieces of structured information. With the addition of `JSON mode` to the model, it is able to output structured information in a reliable manner as detailed in @openai2024text.

@@

```{r}
#| label: tbl-wiki
#| tbl-cap: "Wikidata information about Swedish inventor Jonas Wenström, inventor of three-phase current"

library(tibble)

@@

wenstrom_data %>% 
gt() %>% 
cols_label(key = "Key", value = "Value")

@@

### Prompting and context

Due to the large amount of material that GPT-3.5-turbo has been exposed to in training, it is familiar with the kinds of biographical text that I want it to structure. In order to draw on its familiarity with this kind of material, I need to provide it context about the biographical text it is being passed, and ensure that the model returns output in a useful way. I explain these prompts below.

A "system prompt" can be used to tell the model what kind of material it will be passed, and how it should respond. OpenAI suggest that users "ask the model to adopt a persona" in order to improve responses in a specific task in a guide on prompt engineering [@OpenAI2023PromptEngineering].

I use a simple prompt to tell the model that it will be exposed to Swedish language biographical data:

`system_prompt = "You are an expert on Swedish biographies."`

Next I explain the kinds of information that I want it to extract, and the exact format that I want it in. I do this by specifying a JSON schema, with fields that it must return and specifications for the kind of data in each field. An except of the schema is shown in @fig-schema.

```{r}
#| label: fig-schema
#| fig-cap: "Excerpt of structuring schema"
#| fig-width: 3
#| out-width: 3in

knitr::include_graphics("assets/schema-2.png")
```

Finally I provide detailed context about the source and instructions for what I want the system to do. I include examples of the abbreviations and contractions that it will encounter, and inform the model as to what kind of output I am expecting in return.

`structure_prompt = "Task: read the schema and return RFC compliant JSON information about the Swedish individuals from the 1950 biographical dictionary 'Vem är Vem' that is provided below. Use a numeric index for each biography in your JSON output and return information about all of them, including all career information available. Keep the biographic descriptions in Swedish and remove any abbreviations based on your knowledge, e.g. 'fil. kand.' is 'filosofie kandidat', and 'Skarab. l.' is 'Skaraborgs Län'. Put years in full based on context. Put dates in dd/mm/yyyy format where possible. If there is no information for a key, leave it out. If there is no information for a required key, put NULL as the value.\nHere is the schema: {schema}.\nHere is the text: {page_text}"`

@@

Following this process of structuring the records into a format with specified keys and values, I augment the data by geocoding locations in order to analyze geographic paths of individuals in the sample, and geographic clusters of firms.

@@

person_info %>% 
gt() %>% 
cols_label(key = "Key", value = "Value")

@@

The next task required to understand where these individuals came from and what kinds of firms they were drawn to involved grouping the firms from the catalogue and then classifying each individual in the biographical dictionaries into a particular occupation.

For both of these tasks, I lent on the tools of text embeddings, and a combination of unsupervised machine learning and advanced language processing techniques. I made use of a text embedding model trained on Swedish language text by the @bert-base-swedish-cased. It is an adaptation of the breakthrough BERT model, introduced by Google Research in 2018 [@google2016bert]. The advantage of this KB Lab model is that it has been trained on a selection of Swedish data, including books, news reports, and internet forums. Hence it is able to score the similarity of Swedish business descriptions and occupational titles.

Text embeddings are effective for clustering because they capture semantic meaning rather than relying on surface-level features like character composition. For example, while "steam engine" and "power station" are different in characters and literal meaning, they are semantically related in the context of industrial machinery and energy production. Text embeddings transform these phrases into numerical vectors that reflect this semantic similarity. When applied to clustering, this means that items with similar meanings, even if their literal expressions differ, are grouped together based on the contextual and conceptual similarities encoded in their embeddings. This capability makes text embeddings particularly powerful for organizing and categorizing text data in a way that aligns with human understanding and interpretation [@jurafsky2023speech].

### Classifying occupations

In @jayes_power_2024, we grouped occuaptions into three categories; direct electricity jobs (e.g. electricians), indirect electricity jobs that could benefit from electric motors (e.g. textile workers), and all other jobs. We made this classification based on the occupational title listed in the 1930 census alone. These titles are frequently used in economic history, along with a schema that classifies each title according to a list of possible titles and occupational descriptions. A widely used example is the Historical International Standard Classification of Occupations, or HISCO, defined originally by @vanLeeuwen2002HISCO. A wealth of mappings have been created that link an occupational string like *civilingenjör* to HISCO code 022, civil engineers, described as:

*Workers in this unit group carry out research and advise on civil engineering problems, design projects and structures such as bridges, dams, docks, roads, airports, railways, waste disposal systems, flood control systems and industrial and other large buildings, and plan, organise and supervise their construction, maintenance and repair.*

While this classification process used to involve a large amount of human hours, several new methods that use machine learning have emerged to make this process more efficient. @DahlVedel2024 use a neural network to classify occupational strings using a supervised machine learning process based on the occupational titles in several waves of Danish census data. @merouani2023innovation uses an unsupervised approach based on text embeddings to classify french occupational titles according to the HISCO schema, using a clever compartmentalization strategy to improve classification when individuals hold multiple titles.

In this paper I take a slightly different approach, making use of the additional data from the occupational trajectories of each individual; where they worked and what position they held at each firm or institution. This additional data helps me go beyond just a single occupational string and classify the sector that each engineer worked in. In many instances, the description of the individual in the biographical data relates to their education, rather than what they have done with it.

For example, while it is clear from Axel Nordell's occupational title, *Civilingenjör, kraftverksdirektör* (civil engineer, power station director), that he worked in a directly electricity related occupation, this is not clear for Gustaf Fredrik Ambjörn, whose title is just *Civilingenjör*, based on the fact that he recieved the degree of *Civilingenjör* from the Royal Institute of Technology (KTH) in Stockholm. When looking at his career trajectory, however, we can see that he began his career working at "Fore River Shipbuilding Co., Quincy, Mass., USA" and ended it working as a "Professor i praktiskt skeppsbyggeri" (professor in practical shipbuilding) at Chalmers Institute of Technology in Gothenburg.

I develop a two step process to classify first an individual's occupation based on their occupational title and then a sector based on their occupational trajectory. In both cases I use an unsupervised machine learning approach that turns the titles and work history from text into vector representations of text, and then cluster these vectors in order to assign a HISCO code and then a sector.

To get an intutive understanding of the clustering process, it is worth showing the relative similarity of a sample of occupations from the HISCO schema. @fig-hisco-cluster shows a scatterplot with some of the different occupations listed in HISCO.

We can see a cluster in the top right hand corner of the plot showing basketry weavers and brush makers adjascent to sewers and embroiders, and weaving-and knitting machine setters and pattern card preparers. Thses occupations belong to HISCO major group of production and related workers, and so are all green in the figure. In the bottom left corner we see bacteriolgists, medical doctors and veterinary assistants. On the far right of the plot, we see university lecturers and special education teachers. The takeaway is not necessarily where the occupations fall in the plot, but rather that similar occupations are close to one another in the vector space, and further from dissimilar occupations.

To create this plot, I used the @bert-base-swedish-cased library to create text embedding vectors for each occupation and description in the HISCO schema. Each vector has 364 dimensions, representing the semantic content of the occupation title and description. To produce the figure, I have used dimensionality reducation to collapse the vector representation down from 364 dimensions to just two so that I could plot them on the x and y axis, which I have done with the Uniform Manifold Approximation and Projection for Dimension Reduction (UMAP) library.

```{r}
#| label: fig-hisco-cluster
#| fig-cap: "Scatter plot showing the relative similarity of occupational titles in the HISCO schema"
#| fig-width: 5
#| out-width: 5in

knitr::include_graphics("assets/hisco_cluster.png")
```

In order to classify the occupations from the *Vem är Vem* source according to the HISCO schema, I use this same projection of HISCO occupations, as well as projecting each occupational title into the vector space (in Swedish). I then do a simple cosine distance calculation to find the closest HISCO code to each occupational title in the vector space, and call that the 'most similar HISCO' to the occupational string. I experimented with a threshold for the cosine distance, and found that setting it such that roughly 85 percent of the sample recieved HISCO codes produced results that appeared reasonable. While somewhat arbitrary, this approach allowed for a contextually relevant classification of occupations, drawing on the advanced language understanding capabilities of the KB BERT model. More work to benchmark and score this process is needed in the future.

In the next step, I extract the career trajectories for each engineer and company general manager. I choose engineers and general managers because the other HISCO occupations that make up my sample do not have sufficient information to differentiate their work into sectors. The other largest HISCO occupations in the *Vem är Vem* source are shown in @fig-hisco-counts, being doctors, university teachers, dentists and ministers.

```{r}
#| label: fig-hisco-counts
#| fig-cap: "Most common HISCO occupations in Vem är Vem"
#| fig-width: 8.5
#| out-width: 8.5in

knitr::include_graphics("assets/hisco_counts.png")
```

I group the terms from the career trajectories for each individual, project them into the vector space using the same NLP library as before and cluster them using the kmeans algorithm, an unsupervised learning method. I label these clusters manually based on their content.

[figure here for common types]

### Clustering firms by geography and business

For grouping firms, I again used the k-means clustering algorithm, to categorize firms into 20 distinct categories. This algorithm works by partitioning the data into k distinct clusters based on features (in this case, text embeddings derived from company descriptions) [@jurafsky2023speech].

@@

@@

[In this map, we show two kinds of firms; firms that make electronic products, and firms that produce power or supply power stations with equipment. We can see that electronic products companies are clustered near the large population centers of Stockholm, Gothenburg and Malmö. There are many power stations in central Southern Sweden that make use of hydropower, and firms near them that supply equipment for these power stations.]

@@

## What does the industrial calendar tell us about industrial firms in Sweden in 1947? {.appendix}
